\documentclass{sig-alternate}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

\usepackage{enumitem}
\usepackage{blindtext}
\usepackage{microtype}
    \DisableLigatures[<,>]{family=tt*}
\usepackage{filecontents}
\usepackage{listings}
    \lstset{frame=tbrl,
        numbers=left,
        numberstyle=\tiny,
        basicstyle=\ttfamily\small
        %aboveskip=10pt,
        %belowskip=10pt,
        %literate={`}{\`}1,
    }
    \lstnewenvironment{io}{\lstset{
        numbers=none,
        frameround=tttt,
        basicstyle=\sffamily,
        columns=fullflexible,
        escapechar=@
    }}{}
\usepackage{tabularx}
\usepackage{xcolor}
    \definecolor{medium-blue}{rgb}{0,0,0.5}
\usepackage{hyperref}
    \hypersetup{
        colorlinks, linkcolor={medium-blue},
        citecolor={medium-blue}, urlcolor={medium-blue}
    }
\usepackage[style=numeric,sorting=none]{biblatex}
    \addbibresource{references.bib}
\usepackage[nameinlink]{cleveref}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{textcomp}

\newcommand{\ic}{\textsc{In\-ter\-cal}}
\newcommand{\ascii}{\textsc{As\-cii}}

\title{Esoteric Programming Languages}
\subtitle{An introduction to Brainfuck, INTERCAL, Befunge, Malbolge, and Shakespeare}
\numberofauthors{1}
\author{\alignauthor Sebastian Morr\\\affaddr{Technical University Braunschweig}\\\email{sebastian@morr.cc}}

\makeatletter
\def\@copyrightspace{\relax}
\makeatother
\pagenumbering{arabic}

\begin{document}
\maketitle

\begin{abstract}
    There is a class of programming languages that are not actually designed to be used for programming. These so-called “esoteric” programming languages have other purposes: To entertain, to be beautiful, or to make a point.
    This paper describes and contrasts five influential, stereotypical, and widely different esoteric programming languages: The minimal Brainfuck, the weird \ic{}, the multi-dimensional Befunge, the hard Malbolge, and the poetic Shakespeare.
\end{abstract}

\section{Introduction}
\label{introduction}

While programming languages are usually designed to be used productively and to be helpful in real-world applications, esoteric programming languages have other goals: They can represent proof-of-concepts, demonstrating how minimal a language syntax can get, while still maintaining universality. They might help to prove mathematical theorems or provide boundaries in complexity analyses. The design of esoteric programming languages can be seen an artistic process, and the resulting languages can be expressions of human intellect, wit, and aesthetic taste. Or they might be created as a kind of competitive sport, a challenge for the language's designer or for its users. Finally, there are joke languages designed to enjoy the authors themselves, the users, or even the readers of the specification. All of these motivations and the languages' unique, unusual properties make it worthwile to study esoteric programming languages in more detail.

The word “esoteric” comes from the ancient Greek \emph{esoterikos}, meaning “belonging to an inner circle”, and originally referred to Pythagoras' secret teachings \cite{harper2014esoteric}. It evolved to mean “mystic”, or “having to do with highly theoretical concepts without obvious practical application”. The first usage of the term “esoteric” in the context of “weird programming languages” was probably on a website called \emph{Esoteric Topics in Computer Programming}, published by Chris Pressey, the inventor of Befunge, around 1997 \cite{pressey2005chris}.

For this paper, we picked five iconic esoteric programming languages, each of which demonstrates a unique property:
\textbf{Brainfuck} attempts to have a \textit{minimal} syntax, which consists only of eight different characters. Nevertheless, it can be shown to be Turing-complete. The authors of \textbf{Intercal} wanted to create a \textit{weird} language, that had as many differences to other languages known at that time as possible. \textbf{Befunge} was among the first \textit{two-dimensional} languages, the user can use directional commands to control the program flow. \textbf{Malbolge} was designed to be as \textit{hard} as possible to use. It uses multiple encryption mechanisms, and it took two years to write the first program. Finally, \textbf{Shakespeare} requires its programs to look like Shakespearean plays, making it a \textit{themed} language. It is not obvious at all that the resulting texts are, in fact, meaningful programs. For each language, we are going describe its origins, history, and today's significance. We will explain how the language works, give a meaningful example and mention some popular implementations and variants. 

All five languages are imperative languages. While there are esoteric languages which follow declarative programming paradigms, most of them are quite new and much less popular. It is also interesting to note that esoteric programming languages hardly ever seem to get out of fashion, because of their often unique, weird features. Whereas “real” multi-purpose languages are sometimes replaced by more modern ones that add new features or make programming easier in some way, esoteric languages are not under this kind of pressure, because they do not attempt to be usable. Instead, they can often be appreciated like a piece of art, much like an oil painting can be admired for hundreds of years.

\section{Preliminaries}

We first define some terms that are relevant when discussing esoteric programming languages.

\subsection{Turing machines}

In his 1937 paper \emph{On Computable Numbers} \cite{turing1937computable}, Alan Turing specified a simple machine that he used to define what “computing” means. It consists of an infinite one-dimensional tape with cells that can hold symbols, a head, which moves over the tape and is able to read and write symbols, and a finite state machine, which specifies what to do when a specific symbol is read. This simple architecture is so powerful that it can simulate all other computer models, and thus is able to compute every computable sequence.
In fact, one can construct a Turing machine that reads a symbolic description of another Turing machine, and some input, and then simulates that other machine. This is what we call a \emph{universal Turing machine}.

\subsection{Turing completeness}

If a system can be used to simulate a universal Turing machine, it is called \emph{Turing complete}. By definition this means that this system is as powerful as the class of Turing machines, meaning they as well can compute any computable sequence.
Turing completeness is an important property of universal programming languages. It is highly probable that all modern general-purpose programming languages are Turing complete, but not all esoteric programming languages are, which makes it interesting to look at this property. To be \emph{really} Turing-complete always would require access to an infinite amount of memory, as the systems need to be able to handle arbitrarily large input. Of course, physical machines cannot provide unlimited memory, which is why this restriction is commonly ignored.

\subsection{Standard programs}

When learning or creating a new programming language, one of the first tasks is often to write a program that outputs text, conventionally the string “\texttt{Hello, world!}”. This phrase was made popular by the first edition of the 1978 book \emph{The C Programming Language} by Brian Kernighan and Dennis Ritchie \cite{kernighan1978c}. Being able to write a \emph{Hello world} program in a language demonstrates that it is possible to compile/interpret a simple program in it and that it can output text.

The second standard program is called \emph{99 Bottles of Beer}, which prints the text of a North American folk song. The song repeats lines like “99 bottles of beer on the wall, 99 bottles of beer. Take one down, pass it around, 98 bottles of beer on the wall...” until there are no more bottles left. The ability to program this demonstrates that the language can handle loops and conditional execution.

Finally, another standard programming task is to write a \emph{quine}, a program which prints out its own source code. Usually, it is considered insufficient to simply open the source file and reading from it; instead, the source code must contain all information to reproduce itself. The term was coined by Douglas Hofstadter \cite{hofstadter1979godel} in the honor of Willard Van Orman Quine, an American philosopher who studied self-referentiality. Later, the term was applied to self-replicating programs. The first known quine appeared in a 1972 article by Paul Bratley and Jean Millo and was written in a variant of \textsc{Algol} \cite{bratley1972computer}.

\section{Brainfuck}
\input{brainfuck.tex}

\section{INTERCAL}
\input{intercal.tex}

\section{Befunge}
\input{befunge.tex}

\section{Malbolge}
\input{malbolge.tex}

\section{Shakespeare}
\input{shakespeare.tex}

\section{Conclusion}

As we have seen, esoteric programming languages are certainly entertaining. Besides that, we hypothesize that they also convey a deeper significance:
Esoteric programming languages act as playgrounds for language designers, who can use them to try out features not commonly found in “real” programming languages, without the pressure of having to create something usable. While there is no evidence that this experimentation actually influenced another major programming language, it seems highly probable that, at some point, some language designer made a more informed decision because of his occupation with esoteric programming languages.
Last but not least, they pose interesting, thought-provoking puzzles to the languages' users, which can sharpen their minds and help them to get a better understanding of the different ways to approach and solve problems.

\defbibnote{prenote}{Note: All URLs have been accessed at 2014--12--17.}
\printbibliography[prenote=prenote]

\end{document}
